---
title: "Project Proposal"
subtitle: "due October 11, 2021 by 11:59 PM "
author: "Team E.T.: Elaijah Lapay and Tahmara Ouedraogo"
date: "October 11, 2021"
output: 
  pdf_document:
    latex_engine: xelatex
    df_print: paged
---

# Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse) 
library(readxl)
library(tinytex)
``` 

# Load Data
```{r load-data, message = F}
STAT_198_Food_Security_Dataset <- 
  read_excel("~/R/Team-E-T/data/STAT 198 Food Security Dataset.xlsx", sheet = " County Projections")
```

# Introduction and Data, including Research Questions

(The introduction should introduce your general research question and your data (where it came from, how it was collected, what are the cases, what are the variables, etc.). Your research questions should be clearly specified. The motivation for your research question should be clear, with citations to relevant literature as appropriate.)

# Glimpse

## STAT 198 Food Security Dataset

|variable                                         |description|
|:----------------                                |:-----------|
|County                                           | North Carolina County each row refers to |
|2019 FI Percent                                  |Percent of County Residents Identified as Food Insecure in 2019|
|2019 Child FI Percent                            |Percent of County Residents under 18 years old Identified as Food Insecure in 2019|
|2021 FI Percentage                               |Percent of County Residents Identified as Food Insecure in 2021|
|2019-2021 FI Percent Change                      |Change in Percent of County Residents Identified as Food Insecure from 2019 to 2021|
|2019-2021 Unemployment Change                    |Change in Percent of County Residents Identified as Unemployed from 2019 to 2021|
|2021 Child FI Percent                            |Percent of County Residents under 18 years old Identified as Food Insecure in 2021|
|2019-2021 Child FI Percent Change                |Change in Percent of County Residents under 18 years old Identified as Food Insecure from 2019 to 2021|
|2019 Cost per Meal                               |Estimated Cost per County of the Cost of a Meal from 2019 to 2021|
|2013 Rural-urban Continuum Code                  |2013 Country Designation of Degree of Urbanization and Adjacency to a Metro Area|
|2013 Urban Influence Code                        |2013 Country Designation of Size of Largest City/Town and Adjacency to a Metro Area|
|Less than HS Diploma, 2015-9, Percent            |2015-9 Estimate of County Residents with Less than a High School Diploma in Education|
|Only HS Diploma, 2015-9, Percent                 |2015-9 Estimate of County Residents with only a High School Diploma in Education|
|Some College Education Completed, 2015-9, Percent|2015-9 Estimate of County Residents with Some College Education Completed|
|College Degree Obtained, 2015-9, Percent         |2015-9 Estimate of County Residents with an Associates, Bachelor's, or other Degree Obtained|

(Please use `glimpse` for your data, uploaded into the data folder, here.)
```{r glimpse_data}
glimpse(STAT_198_Food_Security_Dataset)
view(STAT_198_Food_Security_Dataset)
```
# Data Analysis Plan

(Specify the outcome (response, Y) and predictor (explanatory, X) variables you will use to answer your question, as well as the comparison groups you will use, if applicable. You may include very preliminary exploratory data analysis, including some summary statistics and visualizations, along with some explanation on how they help you learn more about your data. Note the statistical method(s) that you believe will be useful in answering your question(s). 
What results from these specific statistical methods are needed to support your hypothesized answer?)
